---
title: "Project 2 — NYCHA Electricity"
author: "Patricio Romero"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

For Project 2, I reviewed the datasets presented in the Week 5 Discussion 5A posts. These posts were used to identify three different datasets that could be prepared and analyzed independently.

This report focuses on the Electric Consumption and Cost dataset. After identifying the dataset through Discussion 5A, I accessed NYC Open Data and downloaded the original data directly from the official source.

The dataset contains monthly electricity consumption and cost records for New York City Housing Authority developments. It includes information such as development name, borough, location, billing period, electrical meter, electricity consumption, and electricity cost.

## Dataset Selection

The Electric Consumption and Cost dataset was selected as one of the three datasets required for Project 2.

Discussion 5A was used to identify and describe the dataset. The actual data used in this report was downloaded from the official NYC Open Data website. The original CSV file will be preserved, and the tidying, transformation, and analysis will be completed in this Project 2 report.

## Data Source

**Dataset:** Electric Consumption and Cost (2010–Sep 2025)  
**Publisher:** New York City Housing Authority  
**Official source:** NYC Open Data  
**Dataset webpage:** [Electric Consumption and Cost](https://data.cityofnewyork.us/d/jr24-e7cr)

The original CSV file is stored in:

```text
Data /NYCHA_Electric_Consumption_And_Cost.csv
```

## Approach

The purpose of this analysis is to compare electricity consumption across the five New York City boroughs and examine how consumption changes over time. The analysis will also evaluate the relationship between electricity consumption and electricity cost.

The original data will be imported from the CSV file downloaded from NYC Open Data. I will use `dplyr` to select the required variables, standardize column names, convert consumption and cost fields to numeric values, and address missing or inconsistent observations.

The selected data will be organized into a wide-format table by month and borough. I will then use `pivot_longer()` from the `tidyr` package to transform the borough columns into a tidy structure. All analysis and visualizations will use the transformed dataset.

### Business Questions

This analysis will address two questions:

1. How does electricity consumption vary across the five New York City boroughs and over time?
2. What is the relationship between electricity consumption and electricity cost?

### Planned Analysis

The analysis will include:

- Monthly electricity consumption by borough.
- A comparison of total and average electricity consumption among the five boroughs.
- An examination of monthly electricity-consumption trends.
- A comparison between electricity consumption and electricity cost.
- Two or three clearly labeled visualizations created with `ggplot2`.
- A brief interpretation of the principal findings.

### Expected Outcome

The analysis will show how electricity consumption differs among the five New York City boroughs and changes over time. It will also examine whether higher electricity consumption is associated with higher electricity cost.

## Data Loading and Initial Exploration

I first loaded the NYCHA dataset into R to check its size, columns, and data types.

The original CSV file was not changed.

```{r}
# Load the required packages
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

# Import the original dataset
nycha <- read_csv(
  "Data /NYCHA_Electric_Consumption_And_Cost.csv",
  show_col_types = FALSE
)

# Display the number of rows and columns
dim(nycha)

# Display the column names
names(nycha)

# Preview the first five observations
head(nycha, 5)

# Examine the dataset structure
glimpse(nycha)
```

### Initial Dataset Summary

The original NYCHA dataset contains 553,666 observations and 27 variables. These records include information about housing developments, electricity billing periods, electricity consumption, and associated costs.

This check helped identify the columns needed for the analysis.


## Data Cleaning and Preparation

The NYCHA Electricity dataset contains 27 variables and 553,666 observations. For this analysis, I selected the variables related to housing developments, boroughs, billing periods, electricity consumption, and electricity costs.

I renamed the selected columns, checked for missing values, and prepared the numbers for analysis.

The original CSV file remains unchanged.

```{r}
# Select and rename the variables needed for the analysis
nycha_clean <- nycha %>%
  transmute(
    development_name = `Development Name`,
    borough = Borough,
    revenue_month = `Revenue Month`,
    bill_id = `UMIS BILL ID`,
    consumption_kwh = readr::parse_number(
      as.character(`Consumption (KWH)`)
    ),
    current_charges = readr::parse_number(
      as.character(`Current Charges`)
    ),
    kwh_charges = readr::parse_number(
      as.character(`KWH Charges`)
    )
  )

# Examine the cleaned dataset
glimpse(nycha_clean)

# Display the first five observations
head(nycha_clean, 5)
```

### Missing Values Assessment

I checked the selected columns for missing values.

```{r}
# Count missing values for each variable
missing_summary <- nycha_clean %>%
  summarise(
    across(
      everything(),
      ~sum(is.na(.))
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "missing_count"
  )

# Display missing values
knitr::kable(
  missing_summary,
  col.names = c("Variable", "Missing Values"),
  caption = "Missing Values by Variable"
)
```

### Borough Validation

I checked the borough names before selecting the five NYC boroughs.

```{r}
# Count records by borough
borough_summary <- nycha_clean %>%
  count(borough, sort = TRUE)

knitr::kable(
  borough_summary,
  col.names = c("Borough", "Number of Records"),
  caption = "Number of Records by Borough"
)
```

### Preparing the Analytical Dataset

I kept records from the five NYC boroughs with valid electricity use and cost values.

The original data was kept unchanged.

```{r}
# Prepare the analytical dataset
nycha_analysis <- nycha_clean %>%
  mutate(
    borough = toupper(trimws(borough))
  ) %>%
  filter(
    borough %in% c(
      "BRONX",
      "BROOKLYN",
      "MANHATTAN",
      "QUEENS",
      "STATEN ISLAND"
    ),
    !is.na(consumption_kwh),
    !is.na(current_charges),
    is.finite(consumption_kwh),
    is.finite(current_charges),
    consumption_kwh >= 0
  )

# Compare original and retained observations
cleaning_summary <- tibble(
  dataset = c("Selected Data", "Analytical Data"),
  observations = c(
    nrow(nycha_clean),
    nrow(nycha_analysis)
  )
)

knitr::kable(
  cleaning_summary,
  col.names = c("Dataset", "Records"),
  caption = "Data Cleaning Summary"
)
```

### Cleaning Summary

After cleaning, 547,893 records remained for analysis.

I used these records to compare electricity use and cost.

## Data Transformation: Wide to Long

I organized the monthly electricity data into wide and long formats.

In the wide table, each borough has its own column.

I used `pivot_longer()` to put the borough names and electricity values into two columns.

The long table is easier to use for charts.

### Monthly Electricity Consumption

```{r}
# Calculate monthly electricity consumption by borough
monthly_consumption <- nycha_analysis %>%
  group_by(revenue_month, borough) %>%
  summarise(
    total_kwh = sum(consumption_kwh, na.rm = TRUE),
    .groups = "drop"
  )

# Display the first observations
knitr::kable(
  head(monthly_consumption, 10),
  col.names = c("Month", "Borough", "Electricity (kWh)"),
  caption = "Monthly Electricity Consumption by Borough"
)
```

### Creating the Wide-Format Dataset

```{r}
# Transform the monthly data into wide format
nycha_wide <- monthly_consumption %>%
  pivot_wider(
    names_from = borough,
    values_from = total_kwh
  ) %>%
  arrange(revenue_month)

# Display the wide-format dataset
knitr::kable(
  head(nycha_wide, 10),
  col.names = c("Month", "Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island"),
  caption = "Wide-Format Electricity Consumption Dataset"
)
```

### Converting Wide Data to Long Data

```{r}
# Convert wide-format data into tidy long format
nycha_long <- nycha_wide %>%
  pivot_longer(
    cols = -revenue_month,
    names_to = "borough",
    values_to = "total_kwh"
  ) %>%
  arrange(revenue_month, borough)

# Display the transformed dataset
knitr::kable(
  head(nycha_long, 10),
  col.names = c("Month", "Borough", "Electricity (kWh)"),
  caption = "Long-Format Electricity Consumption Dataset"
)
```

### Transformation Validation

```{r}
# Compare the number of rows
transformation_summary <- tibble(
  dataset = c(
    "Monthly Aggregated Data",
    "Wide-Format Data",
    "Long-Format Data"
  ),
  observations = c(
    nrow(monthly_consumption),
    nrow(nycha_wide),
    nrow(nycha_long)
  )
)

knitr::kable(
  transformation_summary,
  col.names = c("Dataset", "Records"),
  caption = "Wide-to-Long Transformation Summary"
)

# Verify that electricity consumption totals are preserved
validation <- tibble(
  original_total_kwh = sum(monthly_consumption$total_kwh),
  transformed_total_kwh = sum(nycha_long$total_kwh, na.rm = TRUE)
) %>%
  mutate(
    difference_kwh = original_total_kwh - transformed_total_kwh
  )

knitr::kable(
  validation,
  col.names = c("Original (kWh)", "After Transformation (kWh)", "Difference (kWh)"),
  caption = "Electricity Consumption Validation"
)
```

### Transformation Summary

Each row in the long table represents one borough in one month.

Both tables contain the same electricity totals.

The validation found a difference of 0 kWh.

## Business Question 1: Electricity Consumption by Borough and Over Time

**Research Question:** How does electricity consumption vary across the five New York City boroughs and over time?

To answer this question, I used the transformed long-format dataset to compare total and average monthly electricity consumption across the five boroughs. I also examined monthly consumption patterns to identify differences over time.

### Total and Average Electricity Consumption by Borough

```{r}

# Calculate total and average monthly consumption by borough
borough_consumption <- nycha_long %>%
  group_by(borough) %>%
  summarise(
    total_consumption_kwh = sum(total_kwh, na.rm = TRUE),
    average_monthly_kwh = mean(total_kwh, na.rm = TRUE),
    months = sum(!is.na(total_kwh)),
    .groups = "drop"
  ) %>%
  arrange(desc(total_consumption_kwh))

# Display results
knitr::kable(
  borough_consumption,
  digits = 2,
  col.names = c("Borough", "Total Electricity (kWh)", "Monthly Average (kWh)", "Months"),
  caption = "Total and Average Monthly Electricity Consumption by Borough"
)

```

### Visualization 1: Total Electricity Consumption by Borough

```{r}
# Create a bar chart comparing electricity consumption
ggplot(
  borough_consumption,
  aes(
    x = reorder(borough, total_consumption_kwh),
    y = total_consumption_kwh / 1e9,
    fill = borough
  )
) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Electricity Use by Borough",
    x = "Borough",
    y = "Electricity (billion kWh)"
  ) +
  theme_minimal()
```

### Monthly Electricity Consumption Trends

The next visualization examines how electricity consumption changes over time. Monthly consumption values are displayed separately for each borough.

```{r}
# Convert revenue month into a date
nycha_trends <- nycha_long %>%
  mutate(
    month_date = as.Date(paste0(revenue_month, "-01"))
  )

# Create a monthly consumption line chart
ggplot(
  nycha_trends,
  aes(
    x = month_date,
    y = total_kwh / 1e6,
    color = borough
  )
) +
  geom_line(linewidth = 0.7, na.rm = TRUE) +
  labs(
    title = "Monthly Electricity Use",
    x = "Year",
    y = "Electricity (million kWh)",
    color = "Borough"
  ) +
  theme_minimal()
```

### Interpretation of Business Question 1

Brooklyn had the highest electricity use, about 5.43 billion kWh. Manhattan was next, followed by the Bronx. Queens and Staten Island used less electricity.

The monthly chart shows how electricity use changed over time. Differences may be related to the number and size of NYCHA buildings in each borough.

## Business Question 2: Electricity Consumption and Electricity Cost

**Research Question:** What is the relationship between electricity consumption and electricity cost?

To answer this question, I examined the relationship between electricity consumption, measured in kilowatt-hours (kWh), and electricity charges, measured in US dollars.

I aggregated the cleaned data by month and borough to compare monthly electricity consumption with monthly electricity costs. This approach helps determine whether higher electricity consumption is associated with higher charges.

### Monthly Electricity Consumption and Cost

```{r}
# Calculate monthly electricity consumption and charges by borough
monthly_cost <- nycha_analysis %>%
  group_by(revenue_month, borough) %>%
  summarise(
    total_consumption_kwh = sum(consumption_kwh, na.rm = TRUE),
    total_cost_usd = sum(current_charges, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(revenue_month, borough)

# Display the first ten observations
knitr::kable(
  head(monthly_cost, 10),
  digits = 2,
  col.names = c("Month", "Borough", "Electricity (kWh)", "Cost (USD)"),
  caption = "Monthly Electricity Consumption and Cost by Borough"
)
```

### Correlation Analysis

Pearson's correlation coefficient is used to measure the strength and direction of the linear relationship between electricity consumption and electricity cost.

A positive correlation indicates that higher consumption tends to be associated with higher electricity charges. However, correlation does not establish causation.

```{r}
# Calculate the correlation between consumption and cost
cost_correlation <- monthly_cost %>%
  summarise(
    correlation = cor(
      total_consumption_kwh,
      total_cost_usd,
      use = "complete.obs"
    )
  )

knitr::kable(
  cost_correlation,
  digits = 4,
  col.names = c("Correlation"),
  caption = "Correlation Between Electricity Consumption and Cost"
)
```

### Visualization 3: Electricity Consumption vs. Electricity Cost

```{r}
# Create a scatter plot comparing consumption and cost
ggplot(
  monthly_cost,
  aes(
    x = total_consumption_kwh / 1e6,
    y = total_cost_usd / 1e6,
    color = borough
  )
) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "black"
  ) +
  labs(
    title = "Electricity Use and Cost",
    x = "Electricity (million kWh)",
    y = "Cost (million USD)",
    color = "Borough"
  ) +
  theme_minimal()
```

### Electricity Cost Summary by Borough

```{r}
# Calculate total electricity charges by borough
borough_cost <- monthly_cost %>%
  group_by(borough) %>%
  summarise(
    total_consumption_kwh = sum(total_consumption_kwh),
    total_charges_usd = sum(total_cost_usd),
    average_monthly_cost_usd = mean(total_cost_usd),
    .groups = "drop"
  ) %>%
  arrange(desc(total_charges_usd))

knitr::kable(
  borough_cost,
  digits = 2,
  col.names = c("Borough", "Total Electricity (kWh)", "Total Cost (USD)", "Monthly Average Cost (USD)"),
  caption = "Electricity Consumption and Cost Summary by Borough"
)
```

### Interpretation of Business Question 2

Electricity use and cost had a strong positive relationship. The correlation was 0.9319, meaning higher electricity use was generally linked to higher costs.

Brooklyn had the highest total electricity cost, about $767.55 million. Its average monthly cost was about $4.65 million.

Electricity bills also depend on utility rates and other charges, so correlation does not prove causation.

## Conclusions

This project studied NYCHA electricity use and cost in New York City. The original dataset had 553,666 records. After cleaning, 547,893 records were used.

Brooklyn had the highest electricity use, followed by Manhattan and the Bronx. Electricity use and cost were strongly related, with a correlation of 0.9319.

This project shows how R can help clean, organize, and understand real-world data.

### Limitations

The analysis does not include building size, occupancy, or weather. These factors may affect electricity use, so the results do not measure building efficiency.

## AI Use

I used ChatGPT to help organize the report, improve my English, and review the R code. I downloaded the original data from NYC Open Data and checked the tables and results.