---
title: "Project 2 — Water Quality"
author: "Patricio Romero"
date: today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

For Project 2, I reviewed the datasets presented in the Week 5 Discussion 5A posts. These posts were used to identify three different datasets that could be prepared and analyzed independently.

This report focuses on the Water Quality dataset. The dataset contains water chemistry measurements that can be used to examine environmental conditions and compare different water quality parameters.

I selected this dataset because water quality is an important environmental topic. The data may help identify differences in chemical measurements across sampling locations.

## Dataset Selection

The Water Quality dataset was selected as the second of the three datasets required for Project 2.

Discussion 5A was used to identify the dataset. The original CSV file will be preserved, and the tidying, transformation, and analysis will be completed as part of Project 2.

The objective is to organize the water chemistry measurements into a tidy structure that will make the data easier to compare and analyze.

## Data Source

**Dataset:** Water Quality — Chemistry Data\
**Original file:** Chemistry_Data_Wide_20250715.csv\
**File format:** CSV\
**Dataset category:** Environmental and Water Quality Data

The original CSV file is stored in:

``` text
Data /Chemistry_Data_Wide_20250715.csv
```

The original publisher and official dataset webpage will be added after verifying the source.

## Approach

The purpose of this analysis is to compare water chemistry measurements across different sampling locations and identify which parameters show the greatest variation.

The original data will be imported from the CSV file. I will use `dplyr` to select the required variables, standardize column names, review missing values, and prepare the data for transformation.

The original wide-format dataset will be reviewed to identify the columns containing water chemistry measurements. I will then use `pivot_longer()` from the `tidyr` package to transform the measurement columns into a tidy structure.

The transformed dataset will organize the measurements into columns representing the measurement parameter and its recorded value. All analysis and visualizations will use the transformed dataset.

### Business Questions

This analysis will address two questions:

1.  How do water chemistry measurements vary across different sampling locations?
2.  Which water chemistry parameters show the greatest variation?

### Planned Analysis

The analysis will include:

- A review of the available water chemistry parameters.
- A comparison of water chemistry measurements across sampling locations.
- A calculation of summary statistics for selected chemistry parameters.
- An examination of which parameters show the greatest variation.
- Two or three clearly labeled visualizations created with `ggplot2`.
- A brief interpretation of the principal findings.

### Expected Outcome

The analysis will show how water chemistry measurements differ across the available sampling locations and identify which parameters have the greatest variation.

It will also demonstrate how transforming a wide-format dataset into a tidy structure makes environmental data easier to organize, compare, and analyze.

## Data Loading and Initial Exploration

The first step is to load the Water Quality dataset and review its structure.

This helps identify the available columns, water chemistry measurements, and missing values.

```{r}
# Load the required packages
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)

# Import the original dataset
water <- read_csv(
  "Data /Chemistry_Data_Wide_20250715.csv",
  show_col_types = FALSE
)

# Check the number of rows and columns
dim(water)

# Display column names
names(water)

# Preview the data
head(water, 5)

# Check data types
glimpse(water)
```

### Initial Dataset Summary

The dataset contains water chemistry measurements.

The next step is to identify the sampling locations and the chemistry parameters needed for the analysis.

## Missing Values Analysis

Some measurements are missing. The table shows the number of available and missing values for the five selected parameters.

```{r}
# Review missing measurements before transformation
missing_summary <- water %>%
  select(pH_Lab, TN_mgL, NH3_mgL, TotDissN_mgL, TP_ugL) %>%
  summarise(across(everything(),
                   list(Missing = ~sum(is.na(.x)),
                        Available = ~sum(!is.na(.x))))) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("Parameter", ".value"),
    names_pattern = "^(.*)_(Missing|Available)$"
  ) %>%
  mutate(Missing_Percent = round(100 * Missing / (Missing + Available), 1))

knitr::kable(
  missing_summary,
  col.names = c("Parameter", "Missing Values", "Available Values", "Missing (%)"),
  caption = "Missing Values in Selected Water Chemistry Parameters"
)
```

Missing measurements were not replaced with zero. The analysis uses available recorded values only. Some laboratory flags indicate values below reporting limits, which should be interpreted carefully.

## Data Cleaning and Transformation

The data was cleaned and transformed from wide to long format. Missing measurements were excluded from the analysis.

```{r}
# Select water quality measurements
water_clean <- water %>%
  select(
    SiteName, SiteType, EventDate,
    pH_Lab, TN_mgL, NH3_mgL,
    TotDissN_mgL, TP_ugL
  )

# Transform wide data into long format
water_long <- water_clean %>%
  pivot_longer(
    cols = c(
      pH_Lab, TN_mgL, NH3_mgL,
      TotDissN_mgL, TP_ugL
    ),
    names_to = "Parameter",
    values_to = "Value"
  ) %>%
  filter(!is.na(Value))

# Summary of the transformation
transformation_summary <- tibble(
  Step = c("Original Rows", "Original Columns",
           "Long Format Measurements"),
  Count = c(nrow(water), ncol(water), nrow(water_long))
)

knitr::kable(
  transformation_summary,
  col.names = c("Step", "Count"),
  caption = "Data Transformation Summary"
)
```

## Transformation Validation

The transformation was checked to confirm that the available measurements were preserved.

```{r}
# Count measurements before transformation
original_count <- water_clean %>%
  select(
    pH_Lab, TN_mgL, NH3_mgL,
    TotDissN_mgL, TP_ugL
  ) %>%
  summarise(across(everything(), ~sum(!is.na(.x)))) %>%
  unlist(use.names = FALSE) %>%
  sum()

# Count measurements after transformation
long_count <- nrow(water_long)

validation <- tibble(
  Dataset = c("Original Wide Data", "Transformed Long Data"),
  Measurements = c(original_count, long_count)
)

knitr::kable(
  validation,
  col.names = c("Dataset", "Measurements"),
  caption = "Transformation Validation"
)

stopifnot(original_count == long_count)
```

Both datasets contain the same number of available measurements. This confirms that no measurements were lost during transformation.

## Summary Statistics

This table summarizes the available water chemistry measurements.

```{r}
# Calculate summary statistics
water_summary <- water_long %>%
  group_by(Parameter) %>%
  summarise(
    Samples = n(),
    Average = mean(Value),
    Minimum = min(Value),
    Maximum = max(Value),
    .groups = "drop"
  )

knitr::kable(
  water_summary,
  col.names = c(
    "Parameter",
    "Samples",
    "Average",
    "Minimum",
    "Maximum"
  ),
  digits = 2,
  caption = "Water Quality Summary Statistics"
)
```

## Water Quality Analysis

### Question 1: How do water chemistry measurements vary across sampling locations?

The following analysis compares average nitrogen levels across sampling locations.

```{r}
# Average nitrogen by sampling location
nitrogen_by_site <- water_long %>%
  filter(Parameter == "TN_mgL") %>%
  group_by(SiteName) %>%
  summarise(
    Samples = n(),
    Average_Nitrogen = mean(Value),
    .groups = "drop"
  ) %>%
  arrange(desc(Average_Nitrogen))

knitr::kable(
  head(nitrogen_by_site, 10),
  col.names = c(
    "Sampling Location",
    "Samples",
    "Average Nitrogen (mg/L)"
  ),
  digits = 3,
  caption = "Top 10 Locations by Average Nitrogen"
)
```

```{r}
# Plot average nitrogen at the top 10 locations
ggplot(
  head(nitrogen_by_site, 10),
  aes(
    x = reorder(SiteName, Average_Nitrogen),
    y = Average_Nitrogen
  )
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Average Nitrogen by Sampling Location",
    x = "Sampling Location",
    y = "Average Nitrogen (mg/L)"
  ) +
  theme_minimal()
```

Average nitrogen levels differ across sampling locations. The table and chart show the locations with the highest recorded averages.

### Question 2: Which water chemistry parameters show the greatest variation?

The analysis uses the coefficient of variation to compare selected chemical concentrations measured on different scales. pH is excluded because it uses a logarithmic scale.

```{r}
# Calculate variation by parameter
parameter_variation <- water_long %>%
  filter(Parameter != "pH_Lab") %>%
  group_by(Parameter) %>%
  summarise(
    Samples = n(),
    Average = mean(Value),
    Standard_Deviation = sd(Value),
    CV_Percent = if (
      is.finite(mean(Value)) &&
      abs(mean(Value)) > .Machine$double.eps
    ) {
      100 * sd(Value) / abs(mean(Value))
    } else {
      NA_real_
    },
    .groups = "drop"
  ) %>%
  arrange(desc(CV_Percent))

knitr::kable(
  parameter_variation,
  col.names = c(
    "Parameter",
    "Samples",
    "Average",
    "Standard Deviation",
    "Variation (%)"
  ),
  digits = 2,
  caption = "Variation in Water Chemistry Measurements"
)
```

```{r}
# Compare relative variation
ggplot(
  filter(parameter_variation, !is.na(CV_Percent)),
  aes(
    x = reorder(Parameter, CV_Percent),
    y = CV_Percent
  )
) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(
    title = "Variation by Water Quality Parameter",
    x = "Water Quality Parameter",
    y = "Coefficient of Variation (%)"
  ) +
  theme_minimal()
```

Parameters with higher variation percentages have greater relative differences among their recorded measurements.

### Water pH Distribution

```{r}
# Distribution of pH measurements
ggplot(
  filter(water_long, Parameter == "pH_Lab"),
  aes(x = Value)
) +
  geom_histogram(
    bins = 25,
    fill = "steelblue",
    color = "white"
  ) +
  labs(
    title = "Distribution of Water pH",
    x = "pH",
    y = "Number of Measurements"
  ) +
  theme_minimal()
```

This chart shows how the recorded pH values are distributed across the dataset.

## Key Findings

The analysis provides two main findings based on the business questions.

**Question 1: Water Quality by Location**

Total nitrogen levels vary across sampling locations. Kroma Kill has the highest average nitrogen level at 4.540 mg/L, followed by Mill Creek Confluence at 4.137 mg/L.

**Question 2: Variation in Water Chemistry**

Total phosphorus shows the greatest relative variation at 262.08%, followed by total nitrogen at 206.44%.

These results are based on the available measurements. Missing values were excluded from the analysis.

## Conclusion

The Water Quality dataset contains measurements from different sampling locations.

The analysis shows how nitrogen levels vary by location and compares the relative variation of selected chemistry parameters.

Transforming the data from wide to long format made the measurements easier to summarize and visualize.

The results describe the available samples and do not establish whether a location meets water quality standards.


## AI Use

ChatGPT was used to help interpret the assignment requirements, organize the planned approach, improve the English writing, and provide guidance for the R code. I will run, review, and verify the data transformations, analysis, visualizations, and final conclusions.
